### 2012-10-08 notes.tex
Partial code for LL bound for common M0.

{\bf Variational approximation}
\begin{align}
\log p(r_{kj}; u_0, \sigma_0^2, M_0) & = \log \int_{\theta_{kj}} \int_{\mu_j} p(\mu_j, \theta_{kj}, r_{kj})d\mu_jd\theta_{kj} \\
& = \log \int_{\theta_{kj}} \int_{\mu_j} p(\mu_j, \theta_{kj}, r_{kj}) \frac{q(\mu_j, \theta_{kj})}{q(\mu_j, \theta_{kj})}  d\mu_jd\theta_{kj}\\
& \geq  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log p(\mu_j, \theta_{kj}, r_{kj}) -  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log q(\mu_j, \theta_{kj}) \\
& = \E_q \left[ \log p(r_{kj}, \mu_j, \theta_{kj}) \right] - \E_q\left[ \log q(\mu_j, \theta_{kj}) \right] = \mathcal{L}
\end{align}

@@

We now specify an averaging(variational) distribution
\begin{align}
q(\mu_j | \gamma_{1j}, \gamma_{2j}) & \thicksim \text{Gaussian}(\gamma_{1j}, \gamma_{2j})\\
q(\theta_{kj} | \alpha, \beta ) & \thicksim \text{Beta}(\alpha, \beta)
\end{align}

We are now able to decompose $\mathcal{L}$ into its constituent parts
\begin{equation}\label{eqn:composite}
\mathcal{L} = \E_q\log p(\mu_j|u_0,\sigma_0^2) + \E_q \log p(\theta_{kj} | \mu_j; M_0) 
+ \E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) - \E_q \log q(\mu) - \E_q \log q(\theta_{kj})
\end{equation}

@@

\begin{align}
\E_q \log p(\mu_j ; u_0, \sigma_0^2) & = -\frac{1}{2}\log (2\pi\sigma_0^2) -\frac{1}{2}\left[ \gamma_{2} + (\gamma_1 - u_0)^2 \right]
\end{align}

\begin{align}
\E_q \log p(\theta | \mu ; M_0) & = \int_\mu q(\mu) \int_\theta q(\theta) \log p(\theta | \mu, M_0)\\
\begin{split}
& = \log \Gamma(M_0) - \E_q \log \Gamma(\mu M_0) - \E_q \log \Gamma((1-\mu)M_0) \\
& \qquad+(u_0M_0-1)\left( \psi(\alpha) - \psi(\alpha + \beta) \right) + ((1-u_0)M_0-1)\left( \psi(\beta) - \psi(\alpha + \beta) \right)
\end{split}
\end{align}
where $\E_q \log \Gamma(\mu M_0)$ is easily computed numerically. When we get to differentiating with respect to the parameters, we can move that inside the integral. The kernel changes, but it can still be computed numerically.

\begin{align}
\E_q \log p(r | \theta, n) & = \int_\theta \\
\begin{split}
& = \log \Gamma(n+1) - \log \Gamma(r+1) - \log \Gamma(n-r+1) \\
& \qquad +r(\psi(\alpha)-\psi(\alpha + \beta)) +(n-r)(\psi(\beta) - \psi(\alpha+\beta))
\end{split}
\end{align}

\begin{align}
\E_q \log p(r | \theta, n) & = \int_\theta \\
\begin{split}
& = \log \Gamma(n+1) - \log \Gamma(r+1) - \log \Gamma(n-r+1) \\
& \qquad +r(\psi(\alpha)-\psi(\alpha + \beta)) +(n-r)(\psi(\beta) - \psi(\alpha+\beta))
\end{split}
\end{align}

\begin{equation}
\E_q \log q(\mu)  = \text{Negative Gaussian Entropy}
\end{equation}

\begin{equation}
\E_q \log q(\theta)  = \text{Negative Beta Entropy}
\end{equation}

Now that \eqref{eqn:composite} has been written in terms of only the variational and model parameters, we can numerically optimize the system with respect to variational and the model parameters and do coordinate ascent. Maximizing with respect to the variational parameters tightens the bound on the data log-likelihood and maximizing with respect to the model parameters maximizes the likelihood.
\end{document}

>>>>

{\bf Variational approximation}
\begin{align}
\log p(r_{kj}; u_0, \sigma_0^2, M_0) & = \log \int_{\theta_{kj}} \int_{\mu_j} p(\mu_j, \theta_{kj}, r_{kj})d\mu_jd\theta_{kj} \\
& = \log \int_{\theta_{kj}} \int_{\mu_j} p(\mu_j, \theta_{kj}, r_{kj}) \frac{q(\mu_j, \theta_{kj})}{q(\mu_j, \theta_{kj})}  d\mu_jd\theta_{kj}\\
& \geq  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log p(\mu_j, \theta_{kj}, r_{kj}) -  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log q(\mu_j, \theta_{kj}) \\
& = \E_q \left[ \log p(r_{kj}, \mu_j, \theta_{kj}) \right] - \E_q\left[ \log q(\mu_j, \theta_{kj}) \right] = \mathcal{L_{kj}}
\end{align}

@@

We now specify an averaging(variational) distribution
\begin{align}
q(\mu_j | \gamma_{1j}, \gamma_{2j}) & \thicksim \text{Gaussian}(\gamma_{1j}, \gamma_{2j})\\
q(\theta_{kj} | \alpha_{kj}, \beta_{kj} ) & \thicksim \text{Beta}(\alpha_{kj}, \beta_{kj})
\end{align}

We are now able to decompose $\mathcal{L}$ into its constituent parts
\begin{multline}\label{eqn:composite}
\mathcal{L} = K\sum_{j=1}^J \E_q\log p(\mu_j|u_0,\sigma_0^2) + \sum_{j=1}^J\sum_{k=1}^K\E_q \log p(\theta_{kj} | \mu_j; M_0) 
+ \sum_{j=1}^J\sum_{k=1}^K\E_q \log p(r_{kj} | \theta_{kj}, n_{kj})\\
- K\sum_{j=1}^J\E_q \log q(\mu) - \sum_{j=1}^J\sum_{k=1}^K \E_q \log q(\theta_{kj})
\end{multline}

@@

\begin{align}
\E_q \log p(\mu_j ; u_0, \sigma_0^2) & = -\frac{1}{2}\log (2\pi\sigma_0^2) -\frac{1}{2\sigma_0^2}\left[ \gamma_{2j} + (\gamma_{1j} - u_0)^2 \right]
\end{align}

\begin{align}
\E_q \log p(\theta_{kj} | \mu_j ; M_0) & = \int_{\mu_j} q(\mu_j) \int_{\theta_{kj}} q(\theta_{kj}) \log p(\theta_{kj} | \mu_j, M_0)\\
\begin{split}
& = \log \Gamma(M_0) - \E_q \log \Gamma(\mu_j M_0) - \E_q \log \Gamma((1-\mu_j)M_0) \\
& \qquad+(u_0M_0-1)\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right) + ((1-u_0)M_0-1)\left( \psi(\beta_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right)
\end{split}
\end{align}
where $\E_q \log \Gamma(\mu_j M_0)$ is easily computed numerically. When we get to differentiating with respect to the parameters, we can move that inside the integral. The kernel changes, but it can still be computed numerically.

\begin{align}
\E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) & = \int_{\theta_{kj}} \\
\begin{split}
& = \log \Gamma(n_{kj}+1) - \log \Gamma(r_{kj}+1) - \log \Gamma(n_{kj}-r_{kj}+1) \\
& \qquad +r(\psi(\alpha_{kj})-\psi(\alpha_{kj} + \beta_{kj})) +(n_{kj}-r_{kj})(\psi(\beta_{kj}) - \psi(\alpha_{kj}+\beta_{kj}))
\end{split}
\end{align}

\begin{equation}
\E_q \log q(\mu_j)  = -\frac{1}{2} \log(2\pi e \gamma_{2j})
\end{equation}

\begin{multline}
\E_q \log q(\theta_{kj})  = \log \Gamma(\alpha_{kj} + \beta_{kj}) - \log \Gamma(\alpha_{kj}) - \log \Gamma(\beta_{kj}) \\+(\alpha_{kj}-1)\psi(\alpha_{kj}) +(\beta_{kj} -1)\psi(\beta_{kj}) -(\alpha_{kj}+\beta_{kj}-2)\psi(\alpha_{kj}+\beta_{kj})
\end{multline}

Now that \eqref{eqn:composite} has been written in terms of only the variational and model parameters, we can numerically optimize the system with respect to variational and the model parameters and do coordinate ascent. Maximizing with respect to the variational parameters tightens the bound on the data log-likelihood and maximizing with respect to the model parameters maximizes the likelihood.

{\bf Update Equations}
Isolating the terms of \eqref{eqn:composite} that involve $u_0$ gives
\begin{multline}
\mathcal{L}_{[u_0]} = K\sum_{j=1}^J\left(-\frac{1}{2}(\gamma_{1j}-u_0)^2\right) \\+ \sum_{j=1}^J \sum_{k=1}^Ku_0M_0\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj}+\beta_{kj}) \right) - \sum_{j=1}^J \sum_{k=1}^Ku_0M_0\left( \psi(\beta_{kj}) - \psi(\alpha_{kj}+\beta_{kj}) \right)
\end{multline}

Taking the derivative with respect to $u_0$ gives
\begin{equation}
\frac{\partial \mathcal{L}}{\partial u_0} = K\sum_{j=1}^J\gamma_{1j} - KJu_0 +M_0 \sum_{j=1}^J \sum_{k=1}^K \left( \psi(\alpha_{kj}) - \psi(\beta_{kj}) \right)
\end{equation}

Solving for $u_0$ gives
\begin{equation}
u_0 \leftarrow \frac{1}{J}\sum_{j=1}^J \gamma_{1j} +\frac{M_0}{KJ}\sum_{j=1}^J \sum_{k=1}^K \left( \psi(\alpha_{kj}) - \psi(\beta_{kj}) \right)
\end{equation}

Isolating the terms of \eqref{eqn:composite} that involve $\sigma^2_0$ gives
\begin{equation}
\mathcal{L}_{[\sigma^2_0]} = -\frac{J}{2}\log(2\pi\sigma^2_0) -\frac{1}{2\sigma_0^2}\sum_{j=1}^J\left(\gamma_{2j} + (\gamma_{1j}-u_0)^2 \right)
\end{equation}

Taking the derivative with respect to $\sigma^2_0$ gives
\begin{equation}
\frac{\partial \mathcal{L}}{\partial \sigma^2_0} = -\frac{J}{2\sigma^2_0} +\frac{1}{\sigma^4_0}\sum_{j=1}^J\left(\gamma_{2j} + (\gamma_{1j}-u_0)^2\right)
\end{equation}

Setting the derivative equal to zero and solving for $\sigma^2_0$ gives
\begin{equation}
\sigma^2_0 \leftarrow \frac{1}{J}\sum_{j=1}^J\left(\gamma_{2j} + (\gamma_{1j}-u_0)^2\right)
\end{equation}

\end{document}
